Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-10-31 paper/main.tex
more edits

In many practical optimization problems, the objective function is not
explicitly available and solutions are evaluated by means of expensive
prediction models, simulations or physical experiments. When decision variables
are continuous, the use of Bayesian surrogate-models (e.g., Gaussian processes)
is widespread in
optimization~\citep{JonSchWel98go,ForKea2009surrogate}. Motivated by this
success, there have been attempts at adapting such Bayesian optimization
algorithms to the combinatorial case, a notable example being Combinatorial
Efficient Global Optimization
(CEGO)~\citep{ZaeStoBar2014:ppsn,ZaeStoFriFisNauBar2014}. However, the
ruggedness of combinatorial landscapes, which makes local search particularly
effective in the white-box context, lessens the effectiveness of global
surrogate models~\citep{EriPeaGar2019scalable}. Moreover, surrogate models are
expensive to train and optimize. The additional time required by the Bayesian
optimizer may impose a significant overhead in computation time, specially if
computation time of each function evaluation is measured in ``few'' minutes or
hours rather than days and when ``expensive'' refers to resources or economical
cost rather than wall-clock time. \citet{PerLopStu2015si} recently showed that
ant colony optimization (ACO) %~\citep{DorStu2004:book}
is competitive with CEGO
on a black-box version of the travelling salesman problem under a budget of
$1\,000$ function evaluations.  On the other hand, $1\,000$ evaluations is
still a relatively large budget for CEGO, whereas ACO was not designed for such
short budgets. Finally, we note that learning Budget-constrained Combinatorial is a hot topic for a variety of structures besides permutation~\cite{manchanda2020}.

In this work, we propose and analyze Unbalanced Mallows Model (UMM), a
population-based probabilistic algorithm specifically designed for optimizing
black-box problems over a permutation landscape under a limited budget. There
exists a wide body of research on probabilistic models for permutations, such as Mallows~\cite{gMallows} that have already been 
been used in optimization problems successfully,
with the most prominent examples being estimation of distribution algorithms
(EDAs)~\cite{Ceberio2013}, \cite{Ceberio2015}. ACO may also be considered a type of EDA, since
it builds a probability distribution model from which solutions are sampled
with a bias towards the best solutions evaluated so far. This works fall into at least one of these assumptions: function evaluations are cheap or the function evaluation is a white-box.

The proposed UMM drastically differs from EDAs since EDAs assume that function
evaluations are cheap and they can do many of them. On the contrary, UMM aims
at learning as much as possible with a limited number of function
evaluations. Moreover, unlike EDAs, UMM does not rely on a fixed-size
population of solutions that evolves over ``generations''. Instead, taking
inspiration from Bayesian optimizers, UMM considers a sample of permutations
during the whole execution that is increased by one new permutation at each
iteration.
%
Finally, the balance between intensification and exploration is dynamically
adapted by controlling the variance of the probabilistic Mallows model.

The proposed UMM is based on uBorda~\cite{irurozki2020}, a generalization of
the well-known rank aggregation algorithm Borda~\cite{Borda1781}. Rank
aggregation algorithms summarize in a single permutation the information of a
sample of permutations $S$, and they can be seen as functions to compute the
\textit{mean} of sample $S$. The uBorda algorithm has been proposed to compute
a \textit{weighted mean} of sample $S$. UBorda has been show to return the
correct estimator with high probability on streaming contexts; in this paper,
we show its quality on black-box optimization problems. \ekhine{}{revisa esto
  por favor, Manuel}

\manuel{}{
  \begin{itemize}
  \item Who else has used Mallows model for black-box optimization? Ekhine: Va a mirarlo.
  \item Who else has used unbalanced borda for optimization? Ekhine va a añadir 2 frases.
  \end{itemize}
}

This paper structured as follows. Section~\ref{sec:backgroud} summarizes fundamental concepts about permutation spaces, the uBorda algorithm, black-box combinatorial optimization under a limited budget and CEGO, a Bayesian optimizer for such problems. Section~\ref{sec:umm} presents the UMM algorithm, which is the main contribution of this paper. UMM is analyzed using the experimental setup presented in Section~\ref{sec:setup} and results are presented in Section~\ref{sec:analysis}. We end the paper with overall conclusions and a call for further research on this type of optimization scenario (Section~\ref{sec:conclusions}).

@@

Permutations are defined as bijections of the set $[n]$ of integers onto itself. The set of all permutations of $n$ values is denoted as $S_n$ and has cardinality $n!$. We represent permutations as an ordered vector and, we say that position $i$ contains value $j$ whenever $\sigma(i)=j$. We denote permutations with Greek letters with exception of the identity permutation denoted as $e=(1, 2, 3, \dotsc,n)$. We denote the composition of $\sigma$ and $\pi$ as $\sigma\pi$ and the inverse of $\sigma$ as $\sigma^{-1}$, for which the relation $\sigma\sigma^{-1}=e$ always holds.

@@

\subsection{Unbalanced Borda}\label{sec:uborda}
We introduce now the core for our probabilistic algorithm, the uBorda algorithm. Following the convention of the community, the ranking returned by the uBorda algorithm will be denoted uBorda ranking. The uBorda algorithm is a recent generalization of Borda~\cite{irurozki2020}. In uBorda, we consider that each permutation in the sample $\sigma\in S$ has an associated weight $w(\sigma)$ where $w\colon S_n \to \mathbb{R^{+}}$. Then, the score of item $i \in [n]$ is defined as follows:
%
\begin{equation}\label{eq:uborda_score}
B(i) =  \sum_{t\in S}  w(\sigma_t) \sigma_t(i) \enspace,
\end{equation}
%
and, the same as Borda, orders the items $i \in [n]$ increasingly by their score.

Borda and uBorda are equivalent when $w(\sigma)$ is constant for every $\sigma$. However, in other case uBorda ranking will be closer to the rankings in $S$ with smaller value of $w(\sigma)$. Roughly speaking, it is equivalent to replicating in the sample the $\sigma$ with low values of $w(\sigma$).
Clearly, the choice of  $w(\sigma)$ depends on the relevant property of a ranking in a particular domain. For example, uBorda has been shown to have nice theoretical properties for the online estimation of the parameters of a sample with concept drift~\cite{irurozki2020}.
%
% In this paper, we would like to replicate each permutation $\sigma$ a number of times that is inversely proportional to $f(\sigma)$. Therefore, we will use $w(\sigma)=\rho^{f(\sigma)}$ for a $\rho\in(0,1)$.
Interestingly, uBorda is computed in polynomial time.

@@

Let us assume a black-box ``fitness'' function that must be minimized over a
space of permutations $f\colon S_n \to \mathbb{R}$. Being black-box means that
we can evaluate any candidate permutation $\sigma \in S_n$ to obtain
$f(\sigma)$, however, we do not know anything else about the form of
$f$. Moreover, the evaluation of $f$ is expensive in computation time or
resources and, thus, we can only evaluate a limited budget of candidate
permutations. Here, we will study budgets lower than 400. Such expensive
black-box combinatorial problems arise in diverse contexts, for example,
protein folding~\citep{RomKraArn2012protein} and industrial
production~\citep{FerAlvDiaIglEna2014ants}, where the fitness function may
involve expensive simulations for which no closed-form mathematical description
is available. The black-box formulation and limited budget preclude most of the
optimization techniques that are successful in combinatorial optimization, such
as constructive and perturbation heuristics and local search, which makes this
a particularly challenging scenario, perhaps even more than its continuous
counter-part.

@@

\renewcommand{\subsubsection}{\smallskip{}\noindent\textbf}

@@

The key idea behind UMM is that the probabilistic Mallows model (MM) used in
UMM does not simply represent the current sample $S$. Instead, the model
assigns higher probability values to the permutations with known (or expected)
minimal fitness values. This is achieved by means of the uBorda ranking
(Section~\ref{sec:uborda}), which can be seen as replicating each permutation
$\sigma$ in the sample a number of times inversely proportional to its fitness
$f(\sigma)$, i.e., it behaves as if there were many copies of the best
solutions and few of the bad ones. A property of UMM is that the
computational complexity of working with the original sample (Borda) and with the
transformed sample (uBorda) is the same, i.e., polynomial. The most important steps are
the sampling of a new permutation from $MM(\sigma_0,\theta)$ (line~\ref{line:sample} in Algorithm~\ref{alg:umm}) and the update
(\emph{learning}) of the parameters $\sigma_0$, $\theta$ and $\rho$ (line~\ref{line:uborda} in Algorithm~\ref{alg:umm}).

@@

\subsubsection{Learning step.}
At each iteration we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $\rho \in (0,1)$. %
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$, i.e.,
%
\begin{equation}
r_1 \sum_{\sigma\in S}\rho^{f(\sigma)} =  \sum_{\sigma'\in S'}\rho^{f(\sigma')} \quad\text{where}\quad |S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma)
\end{equation}
%
where $S'$ contains the $r_2|S|$ elements of $S$ with the best corresponding fitness values in $F$. The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification: higher values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas smaller values make the sampling more uniform, thus slowing down the learning.

@@

\subsubsection{Sampling step.}\label{sec:sampling}
Sampling from $MM(\sigma_0,\theta)$  can be done efficiently given the parameters  $\sigma_0$ and $\theta$, as studied by \citet{IruCalLoz2016permallows}, where the computational complexity is $O(n\log n)$. % and $O(n^2)$.\MANUEL{X respectively? more than 2}
%
At each iteration of UMM, $\sigma_0=\hat\sigma_0$ is the permutation obtained in the previous learning step. The scale $\theta$ controls the expected value and variance of the distance $D$ of a random permutation $\sigma\sim MM(\sigma_0, \theta)$ to the location parameter $\sigma_0$. Both  the expected value of $D$ and its variance increase monotonically with the value of $\theta$. In particular, the expected distance $\Edist$ is given by~\cite{gMallows}:
%
\begin{equation}\label{eq:expectation}
\Edist = \frac{n \cdot \exp(-\theta)}{1-\exp(-\theta)} - \sum_{j=1}^{n}  \frac{j\cdot\exp(-j  \theta)}{1-\exp(-j\theta)}
\end{equation}

@@

\subsubsection{CEGO.}
We use the implementation of GECO provided by the
authors.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, we use a GA to optimize
the surrogate models with population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA using the model as a surrogate of the true fitness function.  The authors of
CEGO used a budget of $10^4$ surrogate-model evaluations for each run of the
GA. However, with such a setting, a single run of CEGO with a budget of $\FEmax=400$
true fitness evaluations requires \emph{more than a week} of computation time,
which was infeasible for our study. Hence, we set the budget of the GA to
$10^3$. Although a higher GA budget could improve the results of CEGO reported
here, our results are similar (and sometimes better) than the ones reported in
the original papers~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn} due to
running CEGO with $\FEmax=400$ instead of $\FEmax=200$. In fact, we found that the
increase in runtime caused by doubling $\FEmax$ is much smaller than by increasing
the GA budget by a factor of $10$. Nevertheless, we caution that choosing a
different trade-off may lead to different conclusions and more work is needed
to understand CEGO parameters.

\subsubsection{UMM}. UMM uses the same budget of evaluation ($\FEmax$) as CEGO
and initially samples $\minit=10$ solutions uniformly. The other two parameters
of UMM are $r_1$ and $r_2$, which determine the balance between exploration and
exploitation. In the experimental section we will study their impact. We
implemented UMM using Python 3.

\subsubsection{Other settings}.  For simplicity, we use Kendall's $\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all
experiments. Nevertheless,
\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
when using Hamming distance.  We plan to extend UMM to other distance measures in the future.
In all experiments, we consider a maximum budget of $\FEmax=400$ evaluations of
the actual objective function. In a white-box context, state-of-the-art
algorithms for the problems considered here typically evaluate
thousands of solutions, thus, the budget considered
here for the black-box context is extremely limited.

\subsubsection{Benchmark problems.} Experiments with real-world expensive black-box problems would be computationally infeasible. Instead, we consider classical combinatorial optimization problems as black-box optimization problems.

@@

The \emph{Permutation Flowshop Scheduling Problem (PFSP)} is defined by a matrix $[p_{ij}]$ of size $n \times M$ that gives the processing time of a job $i \in [n]$ on a machine $j \in [M]$. All jobs must be processed by all machines in the same order. Given a permutation of the jobs $\sigma$, $C_{i,j}$ denotes the completion time of job at position $\sigma(i)$ on machine $j$, and $C_{\max} = C_{n,M}$  is the completion time of the last job on the last machine, i.e., the makespan. The objective of the PFSP is to find:
\begin{equation}\label{eq:qap}
  \begin{split}
    \min_{\sigma \in S_n}\quad& f(\sigma) = C_{n,M}\\
    \text{s.t.}\quad & C_{1,j} = 0 \quad j\in [M], \qquad C_{i,1} = 0\quad i \in [n]\\
    &C_{i,j} = p_{\sigma(i),j} + \max\{C_{i-1, j}, C_{i, j-1}\}  \ \  i \in \{2,\dotsc,n\},\ j \in\{2,\dotsc, M\}\\
\end{split}
\end{equation}

\paragraph{Datasets}

For the LOP, we generated synthetic instances as follows: sample 200 permutations from a MM of size $n=20$ and $\theta=-\log(uniformity)$, where  $uniformity=\{.5,.7,.9)\}$, i.e., from peaked distributions to almost uniform distributions. Then, generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}(\sigma(i)<\sigma(j))$.

Moreover, we consider real LOP datasets from the LOPLIB repository.\footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}}
In particular, we consider the following types: IO, Random instances of type A and Random instances of type B.

@@

The first part of the experimental analysis concerns the {analyses} of the learning rate, which corresponds to parameter $\rho$ of uBorda. As stated in Section~\ref{sec:sampling}, $\rho$ is automatically set at each iteration in such a way that the $100r_1\%$ of the samples have the $100r_2\%$ of the weight. In this regard, we tried a few different values of parameters $r_1$ and $r_2$ in all the instances. Only values of $r_1 < r_2$ make sense and $r_1 = r_2 = 0.5$ would produce equal weights in the learning step. Thus, we tried values $r_1 = \{0.1, 0.2, \dotsc, 0.5\}$ and $r_2 = \{0.6,\dotsc, 0.9, 0.99\}$. Figure~\ref{fig:heatmaps} shows the best solution found by each independent run (averaged over 10 runs) for instances: \texttt{N-t1d100.01} and \texttt{N-t1d150.01} for the LOP, \texttt{reC05} for the PFSP, \texttt{kra32} for the QAP. This experiment for all the real datasets can be bound in the supplemental material.

>>>>

In many practical optimization problems, the objective function is not
explicitly available and solutions are evaluated by means of expensive
prediction models, simulations or physical experiments. When decision variables
are continuous, the use of Bayesian surrogate-models (e.g., Gaussian processes)
is widespread in
optimization~\citep{JonSchWel98go,ForKea2009surrogate}. Motivated by this
success, there have been attempts at adapting such surrogate-based optimization
algorithms to the combinatorial case~\cite{MorKat2011:evo}, a notable example being Combinatorial
Efficient Global Optimization
(CEGO)~\citep{ZaeStoBar2014:ppsn,ZaeStoFriFisNauBar2014}. However, the
ruggedness of combinatorial landscapes, which makes local search particularly
effective in the white-box context, lessens the effectiveness of global
surrogate models~\citep{EriPeaGar2019scalable}. Moreover, surrogate models are
expensive to train and optimize. The additional time required by the Bayesian
optimizer may impose a significant overhead in computation time, specially if
computation time of each function evaluation is measured in ``few'' minutes or
hours rather than days and when ``expensive'' refers to resources or economical
cost rather than wall-clock time. \citet{PerLopStu2015si} recently showed that
ant colony optimization (ACO) %~\citep{DorStu2004:book}
is competitive with CEGO
on a black-box version of the travelling salesman problem under a budget of
$1\,000$ function evaluations.  On the other hand, $1\,000$ evaluations is
still a relatively large budget for CEGO, whereas ACO was not designed for such
short budgets.
%Finally, we note that learning Budget-constrained Combinatorial is a hot topic for a variety of structures besides permutation~\cite{manchanda2020}.

In this work, we propose and analyze Unbalanced Mallows Model (UMM), a
population-based probabilistic algorithm specifically designed for optimizing
black-box problems over a permutation landscape under a limited budget. There
exists a wide body of research on probabilistic models for permutations, such
as Mallows~\cite{FliVer1986}, that have already been been used in optimization
problems successfully, with the most prominent examples being estimation of
distribution algorithms
(EDAs)~\cite{CebIruMen2014eda,CebMenLoz2015mallows}. ACO may also be considered
a type of EDA, since it builds a probability distribution model from which
solutions are sampled with a bias towards the best solutions evaluated so
far. These classical EDAs  assume that function
evaluations are cheap and/or the fitness function is white-box.\MANUEL{Do EDAs really assume that the fitness function is white-box? ACO doesn't.}
%The proposed UMM drastically differs from the above algorithms.

In contrast, UMM aims at learning as much as possible with a limited number of
function evaluations. Moreover, unlike EDAs, UMM does not rely on a fixed-size
population of solutions that evolves over ``generations''. Instead, taking
inspiration from Bayesian optimizers, UMM considers a sample of permutations
during the whole execution that is increased by one new permutation at each
iteration. The sample is used to learn a Mallows model from which a single new
permutation is sampled. The model is learned by using
uBorda~\cite{IruLobPer2020arxiv}, a generalization of the well-known rank
aggregation algorithm Borda~\cite{Borda1781}. Rank aggregation algorithms
summarize in a single permutation the information of a sample of permutations,
and they can be seen as functions to compute the \emph{sample mean}. The uBorda
algorithm computes a \textit{weighted sample mean} and has been shown to return
the correct estimator with high probability on streaming
contexts~\cite{IruLobPer2020arxiv}.
%
Finally, the balance between intensification and exploration is dynamically
adapted by controlling the variance of the probabilistic Mallows model.

% The proposed UMM is based on uBorda~\cite{IruLobPer2020arxiv}, a generalization
% of the well-known rank aggregation algorithm Borda~\cite{Borda1781}. Rank
% aggregation algorithms summarize in a single permutation the information of a
% sample of permutations, and they can be seen as functions to compute the
% \emph{sample mean}. The uBorda algorithm has been proposed to compute a
% \textit{weighted sample mean} and shown to return the correct estimator with
% high probability on streaming contexts.  in this paper, we show its quality on
% black-box optimization problems.
%\ekhine{}{revisa esto por
%  favor, Manuel}

% \manuel{}{
%   \begin{itemize}
%   \item Who else has used Mallows model for black-box optimization? Ekhine: Va a mirarlo.
%   \item Who else has used unbalanced borda for optimization? Ekhine va a añadir 2 frases.
%   \end{itemize}
% }

This paper structured as follows. Section~\ref{sec:backgroud} summarizes fundamental concepts about permutation spaces, uBorda, black-box combinatorial optimization under a limited budget and CEGO, a Bayesian optimizer for such problems. Section~\ref{sec:umm} presents the UMM algorithm, which is the main contribution of this paper. UMM is analyzed using the experimental setup presented in Section~\ref{sec:setup} and results are presented in Section~\ref{sec:analysis}. We end the paper with overall conclusions and a call for further research on this type of optimization scenario (Section~\ref{sec:conclusions}).

@@

Permutations are defined as bijections of the set $[n]$ of integers onto itself. The set of all permutations of $n$ values is denoted as $S_n$ and has cardinality $n!$. We represent permutations as an ordered vector and, we say that position $i$ contains value $j$ whenever $\sigma(i)=j$. We denote permutations with Greek letters.\MANUEL{Do we use $e$ anywhere? If not, we do not need this definition ``with exception of the identity permutation denoted as $e=(1, 2, 3, \dotsc,n)$.''} We denote the composition of $\sigma$ and $\pi$ as $\sigma\pi$ and the inverse of $\sigma$ as $\sigma^{-1}$, for which the relation $\sigma\sigma^{-1}=(1, 2, 3, \dotsc,n)$ always holds.

@@

\subsection{Unbalanced Borda}\label{sec:uborda}
We introduce now the core for our probabilistic algorithm, the uBorda algorithm. Following the convention of the community, the ranking returned by the uBorda algorithm will be denoted uBorda ranking. The uBorda algorithm is a recent generalization of Borda~\cite{IruLobPer2020arxiv}. In uBorda, we consider that each permutation in the sample $\sigma\in S$ has an associated weight $w(\sigma)$ where $w\colon S_n \to \mathbb{R^{+}}$. Then, the score of item $i \in [n]$ is defined as follows:
%
\begin{equation}\label{eq:uborda_score}
B(i) =  \sum_{t\in S}  w(\sigma_t) \sigma_t(i) \enspace,
\end{equation}
%
and, the same as Borda, orders the items $i \in [n]$ increasingly by their score.

Borda and uBorda are equivalent when $w(\sigma)$ is constant for every $\sigma$. However, in other case uBorda ranking will be closer to the rankings in $S$ with smaller value of $w(\sigma)$. Roughly speaking, it is equivalent to replicating in the sample the $\sigma$ with low values of $w(\sigma$).
Clearly, the choice of  $w(\sigma)$ depends on the relevant property of a ranking in a particular domain. For example, uBorda has been shown to have nice theoretical properties for the online estimation of the parameters of a sample with concept drift~\cite{IruLobPer2020arxiv}.
%
% In this paper, we would like to replicate each permutation $\sigma$ a number of times that is inversely proportional to $f(\sigma)$. Therefore, we will use $w(\sigma)=\rho^{f(\sigma)}$ for a $\rho\in(0,1)$.
Interestingly, uBorda is computed in polynomial time.

@@

Let us assume a black-box ``fitness'' function that must be minimized over a
space of permutations $f\colon S_n \to \mathbb{R}$. Being black-box means that
we can evaluate any candidate permutation $\sigma \in S_n$ to obtain
$f(\sigma)$, however, we do not know anything else about the form of
$f$. Moreover, the evaluation of $f$ is expensive in computation time or
resources and, thus, we can only evaluate a limited budget $\FEmax$ of candidate
permutations. Here, we will study budgets lower than 400. Such expensive
black-box combinatorial problems arise in diverse contexts, for example,
protein folding~\citep{RomKraArn2012protein} and industrial
production~\citep{FerAlvDiaIglEna2014ants}, where the fitness function may
involve expensive simulations for which no closed-form mathematical description
is available. The black-box formulation and limited budget preclude most of the
optimization techniques that are successful in combinatorial optimization, such
as constructive  heuristics and local search, leading to
a particularly challenging scenario, perhaps even more challenging than its continuous
counter-part.

@@

\renewcommand{\subsubsection}[1]{\smallskip{}\noindent\textbf{#1}.}

@@

The key idea behind UMM is that the probabilistic Mallows model (MM) used in
UMM does not simply represent the current sample $S$. Instead, the model
assigns higher probability values to the permutations with known (or expected)
minimal fitness values. This is achieved by means of the uBorda ranking
(Section~\ref{sec:uborda}), which can be seen as replicating each permutation
$\sigma \in S$ a number of times inversely proportional to its fitness
$f(\sigma)$, i.e., it behaves as if there were many copies of the best
solutions and few of the bad ones. A property of UMM is that the
computational complexity of working with the original sample (Borda) and with the
transformed sample (uBorda) is the same, i.e., polynomial. The most important steps are the update
(\emph{learning}) of the parameters $\sigma_0$, $\theta$ and $\rho$ (line~\ref{line:uborda} in Algorithm~\ref{alg:umm}) and
the sampling of a new permutation from $MM(\sigma_0,\theta)$ (line~\ref{line:sample}). % in Algorithm~\ref{alg:umm}).

@@

\subsubsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $\rho \in (0,1)$. %
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$, i.e.,
%
\begin{equation}
  \begin{split}
  r_1 \sum_{\sigma\in S}\rho^{f(\sigma)} =  \sum_{\sigma'\in S'}\rho^{f(\sigma')} \quad\text{where}\; &|S'| = r_2|S| \\[-1.2em]
  &\land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma)
  \end{split}
\end{equation}
%
i.e., $S'$ contains the $r_2|S|$ elements of $S$ with the best corresponding fitness values in $F$. The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification: higher values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas smaller values make the sampling more uniform, thus slowing down the learning.

@@

\subsubsection{Sampling step}\label{sec:sampling}
%
Sampling from $MM(\sigma_0,\theta)$  can be done efficiently given the parameters  $\sigma_0$ and $\theta$, as studied by \citet{IruCalLoz2016permallows}, where the computational complexity is $O(n\log n)$. % and $O(n^2)$.\MANUEL{X respectively? more than 2}
%
At each iteration of UMM, $\sigma_0=\hat\sigma_0$ is the permutation obtained in the previous learning step. The scale $\theta$ controls the expected value and variance of the distance $D$ of a random permutation $\sigma\sim MM(\sigma_0, \theta)$ to the location parameter $\sigma_0$. Both  the expected value of $D$ and its variance increase monotonically with the value of $\theta$. In particular, the expected distance $\Edist$ is given by~\cite{FliVer1986}:
%
\begin{equation}\label{eq:expectation}
\Edist = \frac{n \cdot \exp(-\theta)}{1-\exp(-\theta)} - \sum_{j=1}^{n}  \frac{j\cdot\exp(-j  \theta)}{1-\exp(-j\theta)}
\end{equation}

@@

\subsubsection{CEGO}
%
We use the implementation of GECO provided by the
authors.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, we use a GA to optimize
the surrogate models with population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA using the model as a surrogate of the true fitness function.  The authors of
CEGO used a budget of $10^4$ surrogate-model evaluations for each run of the
GA. However, with such a setting, a single run of CEGO with a budget of $\FEmax=400$
true fitness evaluations requires \emph{more than a week} of computation time,
which was infeasible for our study. Hence, we set the budget of the GA to
$10^3$. Although a higher GA budget could improve the results of CEGO reported
here, our results are similar (and sometimes better) than the ones reported in
the original papers~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn} due to
running CEGO with $\FEmax=400$ instead of $\FEmax=200$. In fact, we found that the
increase in runtime caused by doubling $\FEmax$ is much smaller than by increasing
the GA budget by a factor of $10$. Nevertheless, we caution that choosing a
different trade-off may lead to different conclusions and more work is needed
to understand CEGO parameters.

\subsubsection{UMM}
%
UMM uses the same budget of evaluations ($\FEmax$) as CEGO
and initially samples $\minit=10$ solutions uniformly. The other two parameters
of UMM are $r_1$ and $r_2$, which determine the balance between exploration and
exploitation. In the experimental section we will study their impact. implemented UMM using Python 3.

\subsubsection{Other settings}
%
For simplicity, we use Kendall's $\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all experiments. Nevertheless,
\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
when using Hamming distance and the same may apply to UMM.
We plan to extend UMM to other distance measures in the future.  In all
experiments, we consider a maximum budget of $\FEmax=400$ evaluations of the
actual objective function. In a white-box context, state-of-the-art algorithms
for the problems considered here typically evaluate thousands of solutions,
thus, the budget considered here for the black-box context is extremely
limited.

\subsubsection{Benchmark problems}
%
Experiments with real-world expensive black-box problems would be computationally infeasible. Instead, we consider classical combinatorial optimization problems as black-box optimization problems.

@@

The \emph{Permutation Flowshop Scheduling Problem (PFSP)} is defined by a matrix $[p_{ij}]$ of size $n \times M$ that gives the processing time of a job $i \in [n]$ on a machine $j \in [M]$. All jobs must be processed by all machines in the same order. Given a permutation of the jobs $\sigma$, $C_{i,j}$ denotes the completion time of job at position $\sigma(i)$ on machine $j$, and $C_{\max} = C_{n,M}$  is the completion time of the last job on the last machine, i.e., the makespan. The objective of the PFSP is to find:
\begin{equation}\label{eq:pfsp}
  \begin{split}
    \min_{\sigma \in S_n}\quad& f(\sigma) = C_{n,M}\\
    \text{s.t.}\quad & C_{1,j} = 0 \quad j\in [M], \qquad C_{i,1} = 0\quad i \in [n]\\
    &C_{i,j} = p_{\sigma(i),j} + \max\{C_{i-1, j}, C_{i, j-1}\}  \ \  i \in \{2,\dotsc,n\},\ j \in\{2,\dotsc, M\}\\
\end{split}
\end{equation}

\subsubsection{Datasets}
%
For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbf{I}\{\sigma(i)<\sigma(j)\}$, where $\mathbf{I}\colon S \times S \to \{0,1\}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}

Moreover, we consider real LOP datasets from the LOPLIB
repository.\footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}} In
particular, we consider the following types: IO, Random instances of type A and
Random instances of type B of various sizes $n \in \{50, 100, 150, 200,
250\}$. These are harder problems than the synthetic instances, not only
because of the larger search space, but also because of the different instance
types.

@@

The first part of the experimental analysis concerns the {analyses} of the learning rate, which corresponds to parameter $\rho$ of uBorda. As stated in Section~\ref{sec:sampling}, $\rho$ is automatically set at each iteration in such a way that the $100r_1\%$ of the samples have the $100r_2\%$ of the weight. In this regard, we tried a few different values of parameters $r_1$ and $r_2$ in all the instances. Only values of $r_1 < r_2$ make sense and $r_1 = r_2 = 0.5$ would produce equal weights in the learning step. Thus, we tried values $r_1 = \{0.1, 0.2, \dotsc, 0.5\}$ and $r_2 = \{0.6,\dotsc, 0.9, 0.99\}$. Figure~\ref{fig:heatmaps} shows the best solution found by each independent run (averaged over 10 runs) for instances: \texttt{N-t1d100.01} and \texttt{N-t1d150.01} for the LOP, \texttt{reC05} for the PFSP, \texttt{kra32} for the QAP. We terminated runs that took longer than 6 hours and those combinations are shown as whitespace.  Results for all the datasets can be found in the supplemental material.

\begin{figure}[htb]
  \centering%
  \begin{minipage}{0.49\linewidth}\centering
    \texttt{N-t1d100.01} (LOP)\\[-0.2ex]
    \includegraphics[width=\textwidth]{../img/heatmap_N-t1d100_01}
  \end{minipage}
  \begin{minipage}{0.49\linewidth}\centering
    \texttt{N-t1d50.01} (LOP)\\[-0.2ex]
 \includegraphics[width=\textwidth]{../img/heatmap_N-t1d150_01}
\end{minipage}\\\vskip 1ex
\begin{minipage}{0.49\linewidth}
  \centering
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/heatmap_rec05}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \centering
  \texttt{kra32} (QAP)\\[-0.5ex]
\includegraphics[width=\textwidth]{../img/heatmap_kra32_dat}
\end{minipage}
\caption{Mean fitness (over 10 runs) of different configurations of $r_1$ and $r_2$ (darker is better). Whitespace denotes missing data (runs that took longer than 6 hours).\label{fig:heatmaps}}
\end{figure}
